### 2019-03-21 response/response-demography-1.tex

I have completely rewritten the introduction to better situate the paper in the literature
and to highlight the contributions of the paper.
I now cover the research on using spacing as a measure of son preference, the 
relationship between spacing and child health (both older and younger child), the 
effect on women's health and other outcomes, and, finally, the limited and mixed results
on what determines birth spacing behavior in developing countries compared to developed 
countries.
I use these to motivate why examining how birth spacing has been affected by the use of
sex selection is important, and, in addition, highlight the three main questions I address 
in the paper: the effect of sex selection on birth intervals, the effect of sex selection
on the distribution of births within spells, and, finally, how the effects have varied
with geographical region.

\item Second, a conceptual framework or theoretical argument is needed to
motivate the decision to examine birth intervals and sex composition by
maternal education. Does maternal education proxy for access to sex
selective technology, preferences for sons, etc.? 
In addition, how does
selection into the highest/lowest educated categories change over the 45
years of the study? How could this impact the interpretation of results?

[Response:]

I have substantially expanded the discussion of the choice of non-proportionality 
variables in the ``Explanatory Variables'' section (and added a footnote to that effect 
in the ``Estimation Strategy'' section).
This now covers education, sex composition, and area of residence in more detail.
If there is a more appropriate place for this discussion, I am open to moving it.

When it comes to selection, there are two possible explanations: one economic and one 
cultural.
In the ``economic'' explanation, the use of sex selection is mainly driven by the lower 
fertility that comes from higher opportunity cost of time and the higher income that allow 
those with higher education to use sex selection.
In that case, there is essentially no selection bias since the use of sex selection is 
driven by factors that are identical across those with the same level of education.

The ``cultural'' explanation is that originally those with higher education where also 
those groups with the strongest son preference (presumably also higher caste).
The expansion of access to education brought other groups/castes into connection with
the higher castes and lead to assimilation of the high caste son preference norms.
Presumably this assimilation would not be complete and those groups newer to higher 
education would therefore have lower son preference and therefore lower use of sex 
selection.
This means that any appearance of lower use of sex selection could be a selection story,
and that the initial group's son preference is just as strong.
The cultural explanation also suggests a reason for why the stated son preference appear 
to decline over time: the ``newer'' entrants have lower initial son preference and, 
therefore, the average son preference for high education women will decline over time.
\citep{bhat03}, \citep{pande07}.

Since I have no way of separating out these two possible explanation with the available
data and set up, I have not included this discussion in the paper itself. 
One possible approach would be to included this as a question for future research in the
Conclusion.
If you think it will make the paper stronger, I am, of course, open to including the
discussion there or anywhere else it might fit better.

### 2019-03-21 paper/sex_selection_spacing-ver2.tex
Respond to Editor's question on proportionality variables, especially education

The explanatory variables in $\mathbf{Z}$ and the interactions between them 
are the non-proportional part of the model, which means that they are
interacted with the baseline hazard:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2),
\end{equation}
where $D_j(t)$ is the piece-wise linear baseline hazard and $\mathbf{Z}_1$ captures sex 
composition of previous children, if any, and $Z_2$ captures area of residence.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally,
but to further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and for different 
time periods.

@@

I divide the explanatory variables into two groups.
The first group consists of variables expected to affect the shape of the hazard function
(the $\mathbf{Z}$ variables): 
mother's education, sex composition of previous children, and area of residence.
I chose these variables because the prior literature shows that they affect 
spacing choices and correlate with sex selection.
Increasing the number of variables interacted with the baseline hazard would
further lower the risk of bias but at the cost of requiring more data to 
precisely estimate.

I divide women into three groups based on education attainment: no
education, one to seven years of education, and eight and more years of education.
The models are estimated separately for each education level to reduce
the potential problem of including other variables as proportional.
Increasing education of mothers correlates closely with lower fertility 
\citep{schultz97}, and 
we should expect lower fertility to lead to more use of sex selection.%
\footnote{
Fathers' education has two opposite predicted effects: the associated higher income
should increase fertility and therefore lower the pressure to use sex selection, but
the higher income also makes the use of sex selection cheaper.
In practice, fathers' education had little effect on the hazards and the use of 
sex-selective abortions, and I, therefore, do not include it.
}

I capture sex composition with dummy variables for the
possible combinations for the specific spell, ignoring the ordering of births.
As an example, for the third spell, three groups are used: Two boys,
one girl and one boy, and two girls.
It is, in principle, possible to estimate the model taking into account
the ordering of children, but this would further lower the power of the
test, by adding one additional group for the third spell and four
additional groups for the fourth spell. 
As discussed, the sex composition of previous children affects both the timing
of births and the use of sex-selective abortions.

Area of residence is a dummy variable for the household living in
an urban area.
The cost of children is higher in urban areas than in rural areas, and 
there is easier access to prenatal sex determination in urban areas.
Both factors are likely to lead to higher use of sex-selective abortions
in urban than in rural areas.

>>>>

The explanatory variables in $\mathbf{Z}$ and the interactions between them 
are the non-proportional part of the model, which means that they are
interacted with the baseline hazard:
\begin{equation}
 \mathbf{Z}_{it} = D_j(t) \times (\mathbf{Z}_1 + Z_2 + \mathbf{Z}_1 \times Z_2),
\end{equation}
where $D_j(t)$ is the piece-wise linear baseline hazard and $\mathbf{Z}_1$ captures sex 
composition of previous children, if any, and $Z_2$ captures area of residence.
The remaining explanatory variables, $\mathbf{X}$, enter proportionally,
but to further minimize any potential bias from assuming proportionality, estimations 
are done separately for different levels of mothers' education and for different 
time periods.%
\footnote{
The choice of which variables to use for non-proportionality is discussed in more detail
in the ``Explanatory Variables'' section.
}

@@

I divide the explanatory variables into two groups.
The first group consists of variables expected to affect the shape of the hazard function
(the $\mathbf{Z}$ variables): 
mother's education, sex composition of previous children, and area of residence.
I chose these variables because the prior literature shows that they affect 
spacing choices and the likelihood of using sex selection.%
\footnote{
Increasing the number of variables interacted with the baseline hazard would
further lower the risk of bias but at the cost of requiring more data to 
precisely estimate.
}

The first variable that is likely to affect the hazard rate in a non-proportional manner
is education.
First, there is evidence, even in the absence of sex selection, that women with
different education levels have different hazard profiles, although the size and direction
of the effect vary across areas and time 
\citep{Tulasidhar1993,Whitworth2002,Bhalotra2008,Kim2010,Soest2018}.
Second, as discussed above, the use of sex selection increases with socioeconomic 
status, especially education, and the use of sex selection is unlikely to be the
same throughout a spell, which means that education cannot enter in a proportional
fashion. \citep{retherford03b,jha06,Portner2015b}.
One potential reason for the correlation between education and sex selection is that
increasing education of mothers correlates closely with lower fertility 
\citep{schultz97}.
This lower fertility leads, in turn, to a higher likelihood of using sex selection
\citep{das_gupta97,Guilmoto2009a,Bongaarts2013,Jayachandran2017}.%
\footnote{
In principle, it is possible that part of the higher likelihood is driven by a stronger
preference for son, independently of desired fertility.
There is, however, evidence that higher education women have shown a decline in
son preference \citep{bhat03,pande07}
}
Another reasons is that having more education is closely associated with higher income and 
wealth, which makes it easier to access illegal services such as prenatal sex 
determination.
It is not possible to separate these two effects here.

I divide women into three groups based on education attainment: no
education, one to seven years of education, and eight and more years of education.
The models are estimated separately for each education level to reduce
the potential problem of including other variables as proportional.%
\footnote{
Fathers' education has two opposite predicted effects: the associated higher income
should increase fertility and therefore lower the pressure to use sex selection, but
the higher income also makes the use of sex selection cheaper.
In practice, fathers' education had little effect on the hazards and the use of 
sex-selective abortions, and I, therefore, do not include it.
}

As discussed, the sex composition of previous children affects both the timing
of births and the use of sex-selective abortions 
\citep{retherford03b,jha06,Bhalotra2008,abrevaya09,Portner2015b,Kumar2016,Soest2018}.
I capture sex composition with dummy variables for the
possible combinations for the specific spell, ignoring the ordering of births.
As an example, for the third spell, three groups are used: Two boys,
one girl and one boy, and two girls.
It is, in principle, possible to estimate the model taking into account
the ordering of children, but this would further lower the power of the
test, by adding one additional group for the third spell and four
additional groups for the fourth spell.

Whether a woman lives in an urban or rural area affect both her fertility level and
her access to sex selection \citep{retherford03b,jha06,Portner2015b}.
Area of residence is a dummy variable for the household living in
an urban area.
The cost of children is higher in urban areas than in rural areas, and 
there is easier access to prenatal sex determination in urban areas.
Both factors are likely to lead to higher use of sex-selective abortions
in urban than in rural areas.
